---
title: "Analysis of Healthcare Access vs Chronic Disease Burdon in the American Southwest"
author: "Teresa McCarrell"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, echo = FALSE}
#Taking a look at the figs
library(pacman)
p_load(tidyverse, janitor, skimr, ggthemes, ggpubr, rstatix, purrr, scales)
theme_set(theme_classic())
analysis_objects <- readRDS( "cdc_analysis_objects.rds")
names(analysis_objects)

```
1. Introduction (~100-150 words)

What this report examines: chronic disease burden and preventive care access across counties in 5 southwestern states
 (CA, AZ, NM, OK, TX).
Why these states: your target job markets / personal relevance (can reuse the honest "why this geography" framing you 
used for IPEDS — just applied to job-market states instead of UC campuses).
The questions the report answers (your finalized four):
What does the distribution of chronic disease burden look like across these states?
Is there a relationship between preventive care access and chronic disease burden?
Does that relationship vary by state?
How do major metro counties compare to their state averages?
Use IR/health-analytics vocabulary where natural: benchmarking, prevalence, disparities.

2. Data & Methods (~60-100 words, same target as before)

Data source: CDC PLACES County Data, 2024 release, filtered to Age-adjusted prevalence (explain briefly why
 age-adjusted — comparability across counties with different age structures).
The 5 measures used and what they capture (briefly).
Note the statistical approach: non-parametric methods (Kruskal-Wallis, Dunn's post-hoc, Spearman correlation) 
chosen because prevalence data doesn't reliably meet normality assumptions.
Note the urban/rural classification: your metro_lookup approach, and the caveat that unmatched counties default
 to "Rural" as a simplification, not a rigorously validated designation.

3. Results — organized by your four questions, each with narrative + figure(s):
```{r, fig.width=8, fig.height=4,fig.cap="Figs", echo = FALSE}
analysis_objects$Fig1
analysis_objects$Fig2
analysis_objects$Fig3
analysis_objects$Fig4
analysis_objects$Fig5
analysis_objects$Table1
analysis_objects$Table2
analysis_objects$Table3
analysis_objects$Table4
analysis_objects$Table5
```
Q1 — Distribution: Fig 1a (histogram) + Fig 1b (boxplot), summary stats table, plus your Kruskal-Wallis + effect size
 + Dunn's results. Lead with the finding: state is associated with large effects on 4 of 5 measures, Oklahoma standing
  out in several.

Q2 — Access vs. burden relationship overall: Fig 2 (hero scatter grid), correlation table. Lead with DIABETES × ACCESS2
 as the strongest finding, and the DIABETES × CHECKUP null result / detection-artifact reasoning as a nuance worth
  calling out.
Q3 — Does it vary by state: Fig 3 (state-by-state faceted grid), noting the small-sample caveat for AZ/NM.
Q4 — Metro vs. state averages: the benchmarking table + dumbbell plot, one clear finding about which metro/state gaps 
stand out most.

4. Conclusions & Recommendations (~100-200 words, matching your IPEDS target)

2-3 concrete, actionable recommendations framed as "strengths to build on" / "areas to watch" — same structure that 
worked before.
Should point toward something a health department, health plan, or program evaluator would actually act on
 (targeted outreach, resource allocation, monitoring priorities).
 Which counties shows the strongest case for targeted intervention, based on the combined disease-burden
  and access picture? 
#(Address whether primary differences are state-based or urban-v-rural based)
