---
title: "Analysis of Healthcare Access vs Chronic Disease Burden in the American Southwest"
author: "Teresa McCarrell"
date: "`r Sys.Date()`"
output: 
  html_document: default
---

```{r setup, echo = FALSE}
library(pacman)
p_load(tidyverse, gt, rmarkdown, knitr)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE)

theme_set(theme_classic())
analysis_objects <- readRDS( "cdc_analysis_objects.rds")
```
### Introduction 

Chronic disease burden (diabetes, high blood pressure, and obesity) and preventive care access 
(health insurance coverage and routine checkups) were examined across counties in five American 
southwestern states (AZ, CA, NM, OK, TX). Data was drawn from the CDC PLACES County Data release, 
which provides model-based prevalence estimates for chronic disease and preventive care measures at
 the county level nationwide, based on 2023 Behavioral Risk Factor Surveillance System (BRFSS) survey data.

This report answers four questions: (1) What does the distribution of chronic disease burden look
 like across counties in these states? (2) Is there a relationship between preventive care access
  and chronic disease burden? (3) How does this access-burden relationship vary between states? 
  (4) How do counties associated with major metropolitan areas compare to their state averages, 
  disentangling state-level differences from urban-versus-rural disparities?

CDC advises that PLACES estimates, as model-based small-area predictions, are not intended for
 evaluating the effects of specific programs or policy interventions. Accordingly, this report
  uses PLACES data as a tool for identifying where disparities in health outcomes and access 
  exist, rather than for evaluating the impact of any past or current intervention.

### Data & Methods 

Data was sourced directly from the CDC PLACES County Data portal, filtered prior to downloading
 to 437 counties from five
 southwestern states (AZ, CA, NM, OK, TX) and to five measures spanning chronic disease burden 
 and preventive care access. One rural Texas county (Loving County, population ~60) was excluded
  due to insufficient survey sample size.

All records reflect age-adjusted prevalence, allowing valid comparison
  across counties with differing age structures. 
   Because prevalence estimates were not normally distributed, non-parametric methods were used 
   throughout: Kruskal-Wallis tests with Dunn's post-hoc pairwise comparisons to assess state-level
    differences, and Spearman's rank correlation to assess the relationship between access and burden measures.

Counties were manually classified as urban or rural based on inclusion in a state's principal
 metropolitan area(s); ambiguous cases were resolved by cross-referencing county-to-metro assignments
  for specific regions (e.g., Dallas-Fort Worth). This classification is coarser than a formal designation 
  (e.g., USDA Rural-Urban Continuum Codes) and should be interpreted as a practical approximation rather
   than an official standard. Urban county counts are notably small in Arizona, New Mexico, and Oklahoma
    (two urban counties each), and comparisons involving these states' urban categories should be read 
    with sample size in mind (Table 1).

```{r, fig.width=8, fig.height=4, echo = FALSE}
analysis_objects$n_counties%>%
gt()%>%
cols_label(StateAbbr = "State",
    UrbanStatus = "Urban/Rural",
    Count = "County Count")%>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))%>%
  tab_source_note(source_note = md("Table 1. County counts by state and urban/rural classification."))
```

### Results 

#### Q1: Distribution of chronic disease burden across states

State was significantly associated with all five measures examined (Supplementary Table 1), though the strength of the
 association varied considerably. Effect sizes ranged from moderate for diabetes (ε² = 0.14) to large 
 for the remaining four measures, with annual checkup rate showing the strongest state-level association
  (ε² = 0.73), followed by high blood pressure (ε² = 0.54), uninsured rate (ε² = 0.49), and obesity 
  (ε² = 0.44) (Supplementary Table 2). Diabetes's comparatively weak association with state suggests that other factors,
   such as age distribution or genetic predisposition of the local demographics, may play a larger role
   in diabetes prevalence than in the other measures studied.

California and Oklahoma emerged as the most consistently divergent pair of states: of the seven state-pairings
 showing the largest effect sizes (|z| > 10), three involved Oklahoma and California directly, spanning 
 annual checkups, high blood pressure, and obesity (Table 2). Oklahoma recorded the highest statewide 
 median prevalence of high blood pressure (37.4%) and obesity (40.4%) among the five states, while also
  recording the highest annual checkup rate (76.2%) — a pattern addressed further in the discussion of 
  access-burden relationships below. California, by contrast, recorded the lowest median prevalence across
   nearly every measure, including diabetes (10.1%), high blood pressure (29.4%), and obesity (30.4%) (Table 3). Figures 1 and 2 illustrate these distributional and state-level patterns directly.

```{r, fig.width=8, fig.height=4,fig.cap="Figure 1. Distributions vary widely by state and measure; annual checkup rates show the clearest separation, with Oklahoma and Texas forming distinct, non-overlapping clusters.", echo = FALSE}
analysis_objects$Fig1
```
```{r, fig.width=8, fig.height=6,fig.cap="Figure 2. State is significantly associated with every measure, with Oklahoma and Texas anchoring the extremes of both chronic disease burden and checkup rates.", echo = FALSE}
analysis_objects$Fig2+
coord_cartesian(clip = "off")
```

```{r, fig.cap = "P-values and magnitude for state-by-state pairings", echo = FALSE}
analysis_objects$Table3 %>%
  gt() %>%
  tab_header(title = "Significant Pairwise State Comparisons") %>%
  tab_source_note(source_note = md("Table 2. P-values and Z-statistics are shown for state-by-state pairwise comparisons.
       Only comparisons with an adjusted *p*-value < 0.05 and an absolute Z-statistic 
       greater than 10 are displayed."))%>%
  cols_label(Short_Question_Text = "Measure", group1 = "State 1", group2 = "State 2",
    statistic = "Z Statistic", p.adj = "Adjusted P-Value") %>%
  cols_hide(columns = c(n1, n2)) %>%
  fmt_number(columns = statistic, decimals = 2) %>%
  fmt_scientific(columns = p.adj, decimals = 2)%>%
  tab_style(style = list(cell_fill(color = "goldenrod3"),
      cell_text(weight = "bold", color = "white")),
      locations = cells_column_labels(everything())) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```
```{r, fig.cap = "Statewide median values for each measure - measures are rows and cols are states.", echo = FALSE}

analysis_objects$Table4 %>%
  gt() %>%
  cols_label(Short_Question_Text = "Measure") %>%
  fmt(
    columns = c(AZ, CA, NM, OK, TX),
    fns = function(x) paste0(sprintf("%.1f", x), "%")) %>%
  tab_header(
    title = "Median County Prevalence by State") %>%
  tab_style(style = list(cell_fill(color = "goldenrod3"),
      cell_text(weight = "bold", color = "white")),
      locations = cells_column_labels(everything())) %>%
  cols_align(align = "center",
    columns = c(AZ, CA, NM, OK, TX)) %>%
  tab_source_note(source_note = md("Table 3. Values represent the median county-level
   age-adjusted prevalence for each measure."))%>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```

#### Q2: Relationship between access and burden across the American Southwest

Preventive care access showed a clear but uneven relationship with chronic disease burden. 
Lack of health insurance was positively correlated with all three burden measures, 
with diabetes showing the strongest relationship observed in the entire analysis (ρ = 0.75, p < .001).
Counties with higher uninsured rates consistently showed higher diabetes prevalence. Obesity
  (ρ = 0.40, p < .001) and high blood pressure (ρ = 0.25, p < .001) showed weaker, 
  though still statistically significant, positive relationships with uninsured rates.

Checkup rate told a more complicated story. High blood pressure (ρ = 0.53, p < .001) and obesity (ρ = 0.39, p < .001) were both positively correlated with checkup rates — counties with more residents visiting a doctor annually showed higher, not lower, prevalence of these conditions. Diabetes showed no significant relationship with checkup rate at all (ρ = −0.05, p = .33). Because these measures rely on self-reported, diagnosed conditions rather than direct clinical measurement, this pattern likely reflects a detection effect: more frequent checkups increase the likelihood that an existing condition is identified and recorded, rather than checkups themselves driving disease prevalence upward. This distinction matters for interpretation — it suggests uninsured rate is a more direct proxy for care access in this dataset, while checkup rate is better understood as a proxy for diagnostic detection. Figure 3 illustrates all six relationships directly.

```{r, fig.width=12, fig.height=4,fig.cap="Figure 3. Uninsured rate shows a consistent positive relationship with chronic disease burden, strongest for diabetes; checkup rate's relationship is inconsistent and, for diabetes, statistically null.", echo = FALSE}
analysis_objects$Fig3+
  theme(strip.text = element_text(size = 8))
```

#### Q3: Does the access-burden relationship vary between states?

The diabetes-uninsured relationship held across every state, though its strength varied meaningfully:
 from ρ = 0.63 in Oklahoma to ρ = 0.89 in Texas, with California (ρ = 0.88), Arizona (ρ = 0.70),
  and New Mexico (ρ = 0.67) falling between. This consistency — a significant, positive relationship
   in all five states — makes it the most reliable finding in the analysis (Figure 4, Table 4). The obesity-uninsured
    relationship was also significant in every state, but with more variability in strength,
     ranging from ρ = 0.29 in Oklahoma to ρ = 0.72 in Texas. The high blood pressure-uninsured
      relationship was directionally consistent across all five states but did not reach significance
       in Arizona (ρ = 0.50, p = .06); potentially a reflection of Arizona's small county count (n = 15)
        rather than a true absence of relationship, given that the direction and magnitude were consistent
        with those of the other sampled states.

The checkup-based relationships were far less consistent by state.
 Texas and California were the only states where checkup rate showed significant relationships with
  burden measures — and in both states, all three were negative (better care-detection proxy,
   lower detected burden), consistent with the overall pattern. In Arizona, New Mexico, and Oklahoma,
    no checkup-burden relationship reached significance for any of the three conditions. This split
     suggests the checkup-detection relationship identified in the overall analysis may be driven
      primarily by Texas and California's larger sample sizes.

```{r, fig.width=14, fig.height=8,fig.cap="Figure 4. The diabetes-uninsured relationship is the only one that holds significantly across all five states; checkup-based relationships reach significance only in Texas and California.", echo = FALSE}
analysis_objects$Fig4+
coord_cartesian(clip = "off")
```
```{r, fig.width=6, fig.height=4, echo = FALSE}
analysis_objects$state_correlation_summary%>%
select(StateAbbr,pair_name,rho,p_value)%>%
filter(p_value < 0.05, abs(rho) >= 0.40)%>%
arrange(StateAbbr, desc(abs(rho)))%>%
group_by(StateAbbr)%>%
gt()%>%
  cols_label(StateAbbr = "State", pair_name = "Measure Pair",
    rho = "Spearman's ρ", p_value = "P-Value") %>%
  fmt_number(columns = rho,
    decimals = 2) %>%
  fmt(columns = p_value,
    fns = function(x) ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))) %>%
  tab_header(title = "Strong Within-State Correlations") %>%
  tab_source_note(source_note = md( "Table 4. Only statistically significant correlations (*p* < 0.05) 
      with an absolute Spearman correlation of at least 0.40 are shown."))%>%
  tab_style(style = list(cell_fill(color = "goldenrod3"),
      cell_text(weight = "bold", color = "white")),
      locations = cells_column_labels(everything())) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```

#### Q4: Metro vs. state averages
Rural counties showed higher uninsured rates than their state's urban counties in four of five states:
California (9.7% rural vs. 8.8% urban), Arizona (14.6% vs. 13.1%), New Mexico (17.5% vs. 14.2%),
 and most notably Texas (19.0% vs. 14.5%, a 4.6 percentage-point gap, the largest urban-rural
  insurance gap observed). Oklahoma was the exception, with rural and urban insurance rates nearly
   identical (12.1% vs. 12.2%).
```{r, fig.width=9, fig.height=6,fig.cap="Figure 5. New Mexico shows the widest rural-urban gaps across nearly every measure, while Oklahoma's gaps stay narrow despite having the region's highest overall burden.", echo= FALSE}
analysis_objects$Fig5
```
This urban-rural insurance gap generally tracked with the disease burden gap in the same direction:
 rural counties also showed higher obesity, diabetes, and high blood pressure prevalence than urban
  counties in nearly every state, with New Mexico showing the widest rural-urban obesity gap observed
   anywhere in the dataset (36.5% rural vs. 29.6% urban, a 6.95 percentage-point difference). Oklahoma
    again stood apart, showing comparatively small urban-rural gaps across most measures despite having
     the highest statewide burden overall — indicating Oklahoma's chronic disease burden is elevated broadly
      across both urban and rural counties, rather than concentrated in one or the other.

Combined with the Q3 findings, this points to New Mexico and Texas as the states where rural-urban disparities
 in access and burden are most pronounced, while Oklahoma's elevated burden appears to be a statewide pattern
  not explained by the urban-rural divide.

However, not all identified gaps were found to be statistically significant. Figure 6 confirms which of these rural-urban gaps
 reach statistical significance. Notably, most of Oklahoma's gaps do not reach significance despite Oklahoma's high
   overall burden. Only in Texas were all five measures significantly different between rural and urban counties.
```{r, fig.width=9, fig.height=9,fig.cap="Figure 6. Texas is the only state where rural-urban differences reach statistical significance across all five measures.", echo= FALSE}
analysis_objects$Fig6+
coord_cartesian(clip = "off")
```

**Note: as flagged in Methods, urban county counts for Arizona, New Mexico, and Oklahoma are small (n = 2 each),
 so these states' urban-category comparisons should be treated as indicative rather than statistically robust.**


### Conclusions & Recommendations

This analysis points to insurance coverage, rather than checkup frequency, as the more reliable lever for
 addressing chronic disease burden in the Southwest. The relationship between uninsured rate and diabetes
  prevalence was not only the strongest observed in this analysis but also the most consistent, holding
   significantly across all five states. This consistency makes expanding insurance access a stronger,
    better-supported priority than checkup-based outreach specifically for diabetes, where the detection-driven
     checkup relationship offered no significant signal at all.

Texas and New Mexico warrant particular attention for rural outreach: both states show the widest gaps
 between rural and urban counties in insurance coverage, and those same states show correspondingly wide
  rural-urban gaps in chronic disease burden. Rural-targeted insurance enrollment efforts may offer the
   clearest opportunity for impact in these two states specifically.

Oklahoma presents a different pattern and requires a different response: its elevated chronic disease burden
 appears broadly across both urban and rural counties rather than concentrated in one, suggesting statewide
  programming may be more appropriate than a rural-urban targeted approach.

As these estimates are modeled prevalence data rather than intervention outcomes, they are best used to
 identify where to prioritize further investigation and resource planning, not as evidence of any specific
  program's effect.


### Appendix

The following tables provide additional statistical detail underlying the tests summarized in the Results section.

```{r,echo = FALSE}
analysis_objects$Table1%>%
  gt()%>%
  cols_label(Short_Question_Text = "Measure")%>%
  tab_source_note(source_note = md("Supplementary Table 1: Table of measure and p-value; for data shown in Figure 2."))%>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```
```{r, echo = FALSE}
analysis_objects$Table2%>%
  select(Short_Question_Text, effsize, magnitude)%>%
  gt() %>%
  cols_label(Short_Question_Text = "Measure", effsize = "Effect Size", magnitude = "Magnitude")%>%
  tab_source_note(source_note = md("Supplementary Table 2:Effect size and magnitude for data shown in Figure 2")) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```
```{r, echo = FALSE}
analysis_objects$Table5%>%
  gt()%>%
  cols_label(Short_Question_Text = "Measure", StateAbbr = "State", diff = "Difference")%>%
  tab_source_note(source_note = md("Supplementary Table 3: Detailed values associated with figure 5")) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```