---
title: "Analysis of Healthcare Access vs Chronic Disease Burden in the American Southwest"
author: "Teresa McCarrell"
date: "`r Sys.Date()`"
output: 
  html_document: default
---

```{r setup, echo = FALSE}
library(pacman)
p_load(tidyverse, gt, rmarkdown, knitr)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE)

theme_set(theme_classic())
analysis_objects <- readRDS( "cdc_analysis_objects.rds")
```
### Introduction 

Chronic disease burden (diabetes, high blood pressure, and obesity) and preventive care access 
(health insurance coverage and routine checkups) were examined across counties in five American 
southwestern states (Arizona, California, New Mexico, Oklahoma, Texas). Data was drawn from the public CDC PLACES County Data release, 
which provides model-based prevalence estimates for chronic disease and preventive care measures at
 the county level nationwide, based on 2023 Behavioral Risk Factor Surveillance System (BRFSS) survey data.

This report answers four questions: (1) What does the distribution of chronic disease burden look
 like across counties in these states? (2) Is there a relationship between preventive care access
  and chronic disease burden? (3) How does this access-burden relationship vary between states? 
  (4) How do counties associated with major metropolitan areas compare to their rural areas within the same state?

 Although this report
  uses PLACES data as a tool for identifying where disparities in health outcomes and access 
  exist, the CDC advises that PLACES estimates, as model-based small-area predictions, are not intended to be the sole 
  data source for
 evaluating the effects of specific programs or determining the targets of policy interventions.

### Data & Methods 

Data was sourced directly from the CDC PLACES County Data portal, filtered prior to downloading. Filters were set
 acquire data for five measures spanning chronic disease burden 
 and preventive care access, and for 437 out of the 438 counties from the five
 southwestern states (Arizona, California, New Mexico, Oklahoma, Texas). 
 One rural Texas county (Loving County, population ~60) was excluded
  due to insufficient survey sample size.

All records reflect age-adjusted prevalence, allowing valid comparison
  across counties with differing age structures. 
Non-parametric methods were used due to non-normal data distributions;
Kruskal-Wallis tests with Dunn's post-hoc pairwise comparisons to determine state-level
    differences, and Spearman's rank correlation to evaluate the relationship between access and burden measures.

Counties were manually classified as urban or rural based on inclusion in a state's principal
 metropolitan area(s). Ambiguous cases were resolved by cross-referencing county-to-metro assignments (e.g., searching for
 which counties are considered part of Dallas-Fort Worth). This classification is coarser than a formal designation 
  (e.g., USDA Rural-Urban Continuum Codes). Urban county counts are notably small in Arizona, New Mexico, and Oklahoma
    (two urban counties each), in comparision to urban county counts in Texas and California. Cmparisons involving 
    urban vs rural counties should be read with sample size in mind (Table 1).

```{r, fig.width=8, fig.height=4, echo = FALSE}
analysis_objects$n_counties%>%
gt()%>%
cols_label(StateAbbr = "State",
    UrbanStatus = "Urban/Rural",
    Count = "County Count")%>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))%>%
  tab_source_note(source_note = md("Table 1. County counts by state and urban/rural classification."))
```

### Results 

#### #1: What is distribution of chronic disease burden across counties in these states?

State was significantly associated with all five measures examined (Supplementary Table 1), though there was cosiderable
variation in the strength of the association. Effect sizes ranged from moderate for diabetes (ε² = 0.14), to large 
 for the remaining four measures. Annual checkup rate showed the strongest state-level association
  (ε² = 0.73), followed by high blood pressure (ε² = 0.54), uninsured rate (ε² = 0.49), and obesity 
  (ε² = 0.44) (Supplementary Table 2). The comparatively weak association of diabetes with state suggests that other factors,
   such as the genetic predisposition of the local demographic, may play a larger role
   in diabetes prevalence than in the other measures studied.

California and Oklahoma emerged as the most consistently divergent pair of states. Of the seven state-measure  pairings
 showing the largest effect sizes (|z| > 10), three involved Oklahoma and California, spanning 
 annual checkups, high blood pressure, and obesity (Table 2). Oklahoma recorded the highest statewide 
 median prevalence of high blood pressure (37.4%) and obesity (40.4%) among the five states, while also
  recording the highest annual checkup rate (76.2%). California, by contrast, recorded the lowest median prevalence across
   nearly every measure, including diabetes (10.1%), high blood pressure (29.4%), and obesity (30.4%) (Table 3). 
   Figures 1 and 2 illustrate the dataset-wide measure distributions and state-level patterns directly.

```{r, fig.width=8, fig.height=4,fig.cap="Figure 1. Distributions vary widely by state and measure; annual checkup rates show the clearest separation, with Oklahoma and Texas forming distinct, non-overlapping clusters.", echo = FALSE}
analysis_objects$Fig1
```
```{r, fig.width=8, fig.height=6,fig.cap="Figure 2. State is significantly associated with every measure, with Oklahoma and Texas anchoring the extremes of both chronic disease burden and checkup rates.", echo = FALSE}
analysis_objects$Fig2+
coord_cartesian(clip = "off")
```

```{r, fig.cap = "P-values and magnitude for state-by-state pairings", echo = FALSE}
analysis_objects$Table3 %>%
  gt() %>%
  tab_header(title = "Significant Pairwise State Comparisons") %>%
  tab_source_note(source_note = md("Table 2. P-values and Z-statistics are shown for state-by-state pairwise comparisons.
       Only comparisons with an adjusted *p*-value < 0.05 and an absolute Z-statistic 
       greater than 10 are displayed."))%>%
  cols_label(Short_Question_Text = "Measure", group1 = "State 1", group2 = "State 2",
    statistic = "Z Statistic", p.adj = "Adjusted P-Value") %>%
  cols_hide(columns = c(n1, n2)) %>%
  fmt_number(columns = statistic, decimals = 2) %>%
  fmt_scientific(columns = p.adj, decimals = 2)%>%
  tab_style(style = list(cell_fill(color = "goldenrod3"),
      cell_text(weight = "bold", color = "white")),
      locations = cells_column_labels(everything())) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```
```{r, fig.cap = "Statewide median values for each measure - measures are rows and cols are states.", echo = FALSE}

analysis_objects$Table4 %>%
  gt() %>%
  cols_label(Short_Question_Text = "Measure") %>%
  fmt(
    columns = c(AZ, CA, NM, OK, TX),
    fns = function(x) paste0(sprintf("%.1f", x), "%")) %>%
  tab_header(
    title = "Median County Prevalence by State") %>%
  tab_style(style = list(cell_fill(color = "goldenrod3"),
      cell_text(weight = "bold", color = "white")),
      locations = cells_column_labels(everything())) %>%
  cols_align(align = "center",
    columns = c(AZ, CA, NM, OK, TX)) %>%
  tab_source_note(source_note = md("Table 3. Values represent the median county-level
   age-adjusted prevalence for each measure."))%>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```

#### #2: Is there a relationship between preventive care access and chronic disease burden?

Preventive care access showed a clear but uneven relationship with chronic disease burden. 
Lack of health insurance was positively correlated with all three burden measures, 
with diabetes showing the strongest relationship in the analysis (ρ = 0.75, p < .001).
Counties with higher uninsured rates consistently showed higher diabetes prevalence. Obesity
  (ρ = 0.40, p < .001) and high blood pressure (ρ = 0.25, p < .001) also showed statistically significant, positive relationships with uninsured rates,
  albiet with weaker correlation.

High blood pressure (ρ = 0.53, p < .001) and obesity (ρ = 0.39, p < .001) were both positively
 correlated with checkup rates. Counties where more residents visited a doctor annually showed higher, not lower, prevalence of these conditions.
  Diabetes showed no significant relationship with checkup rate (ρ = −0.05, p = .33). Because these measures rely on self-reported, diagnosed
   conditions rather than direct clinical measurement, this pattern likely reflects a detection effect: more frequent checkups increase the likelihood
    that an existing condition is identified and recorded. This suggests that uninsured
     rate is a more direct proxy for care access in this dataset, while checkup rate is better understood
      as a proxy for diagnostic detection. Figure 3 illustrates the six relationships directly.

```{r, fig.width=12, fig.height=4,fig.cap="Figure 3. Uninsured rate shows a consistent positive relationship with chronic disease burden, strongest for diabetes; checkup rate's relationship is inconsistent and, for diabetes, statistically null.", echo = FALSE}
analysis_objects$Fig3+
  theme(strip.text = element_text(size = 8))
```

#### #3: How does this access-burden relationship vary between states?

First, the relationship between uninsured rates and various burdens were evaluated within each state.
The diabetes-uninsured relationship was significant and positively correlated across every state, though its strength varied: ranging
 from ρ = 0.63 in Oklahoma to ρ = 0.89 in Texas, with California (ρ = 0.88), Arizona (ρ = 0.70),
  and New Mexico (ρ = 0.67) falling between (Figure 4, Table 4). The obesity-uninsured
    relationship was also significant in every state, with variability in strength,
     ranging from ρ = 0.29 in Oklahoma to ρ = 0.72 in Texas. The high blood pressure-uninsured
      relationship was directionally consistent across all five states but did not reach statistical significance
       in Arizona (ρ = 0.50, p = .06). This nay be due to Arizona's small county count (n = 15)
        rather than a true absence of relationship, given that the direction and magnitude were consistent
        with those of the other sampled states.

Secondly, the relationship between checkup and burdens were evaluated. Checkup-burden relationships were less consistent 
than uninsured-burden relationships.
 Texas and California were the only states where checkup rate showed significant relationships with
  burden measures; and all three relationships were negative, consistent with the overall pattern of better access to detection resulting in
  a higher rate of detection of existing conditions. In Arizona, New Mexico, and Oklahoma,
    no checkup-burden relationship reached significance for any of the three measured burdens. This split
     suggests the checkup-detection relationship identified in the overall analysis may be influenced
      by Texas and California's larger sample sizes.

These results are summarized in Figure 4 below.

```{r, fig.width=15, fig.height=9,fig.cap="Figure 4. The diabetes-uninsured relationship is the only one that holds significantly across all five states; checkup-based relationships reach significance only in Texas and California.", echo = FALSE}
analysis_objects$Fig4+
coord_cartesian(clip = "off")
```
```{r, fig.width=6, fig.height=4, echo = FALSE}
analysis_objects$state_correlation_summary%>%
select(StateAbbr,pair_name,rho,p_value)%>%
filter(p_value < 0.05, abs(rho) >= 0.40)%>%
arrange(StateAbbr, desc(abs(rho)))%>%
group_by(StateAbbr)%>%
gt()%>%
  cols_label(StateAbbr = "State", pair_name = "Measure Pair",
    rho = "Spearman's ρ", p_value = "P-Value") %>%
  fmt_number(columns = rho,
    decimals = 2) %>%
  fmt(columns = p_value,
    fns = function(x) ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))) %>%
  tab_header(title = "Strong Within-State Correlations") %>%
  tab_source_note(source_note = md( "Table 4. Only statistically significant correlations (*p* < 0.05) 
      with an absolute Spearman correlation of at least 0.40 are shown."))%>%
  tab_style(style = list(cell_fill(color = "goldenrod3"),
      cell_text(weight = "bold", color = "white")),
      locations = cells_column_labels(everything())) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```

#### #4: How do counties associated with major metropolitan areas compare to their rural areas within the same state?
Rural counties showed higher uninsured rates than their state's urban counties in four of five states:
California (9.7% rural vs. 8.8% urban), Arizona (14.6% vs. 13.1%), New Mexico (17.5% vs. 14.2%),
 and most notably Texas (19.0% vs. 14.5%, a 4.6 percentage-point gap, the largest urban-rural
  insurance gap observed). Oklahoma was the exception, with rural and urban insurance rates nearly
   identical (12.1% vs. 12.2%) (Figure 5).
```{r, fig.width=9, fig.height=6,fig.cap="Figure 5. New Mexico shows the widest rural-urban gaps across nearly every measure, while Oklahoma's gaps stay narrow despite having the region's highest overall burden.", echo= FALSE}
analysis_objects$Fig5
```
This urban-rural insurance gap generally tracked with the disease burden gap in the same direction:
 rural counties also showed higher obesity, diabetes, and high blood pressure prevalence than urban
  counties in nearly every state, with New Mexico showing the widest rural-urban obesity gap observed
   anywhere in the dataset (36.5% rural vs. 29.6% urban, a 6.95 percentage-point difference). Oklahoma
    again stood apart, showing comparatively small urban-rural gaps across most measures despite having
     the highest statewide burden overall. This indicates that Oklahoma's chronic disease burden is elevated broadly
      across both urban and rural counties, rather than concentrated in one or the other.

These results suggest that New Mexico and Texas are the states where rural-urban disparities
 in access and burden are most pronounced, while Oklahoma's elevated burden is potentially a statewide pattern
  not explained by the urban-rural divide.

However, not all identified gaps were found to be statistically significant. Figure 6 confirms which of these rural-urban gaps
 reach statistical significance. Notably, most of Oklahoma's gaps do not reach significance despite Oklahoma's high
   overall burden. Only in Texas were all five measures significantly different between rural and urban counties.
```{r, fig.width=9, fig.height=9,fig.cap="Figure 6. Texas is the only state where rural-urban differences reach statistical significance across all five measures.", echo= FALSE}
analysis_objects$Fig6+
coord_cartesian(clip = "off")
```

**Note: as flagged in Methods, urban county counts for Arizona, New Mexico, and Oklahoma are small (n = 2 each),
 so these states' urban-category comparisons should be treated as indicative rather than statistically robust.**


### Conclusions & Recommendations

This analysis points to insurance coverage, rather than checkup frequency, as the more reliable lever for
 addressing chronic disease burden in the Southwest. The relationship between uninsured rate and diabetes
  prevalence was not only the strongest observed in this analysis but also the most consistent, being statistically
   significant across all five states. This consistency makes expanding insurance access a stronger,
    better-supported priority than checkup-based outreach, particularly for diabetes, where the detection-driven
     checkup relationship offered no significant signal at all.

Texas and New Mexico warrant particular attention for rural outreach. These states show the widest gaps
 between rural and urban counties in insurance coverage, and show correspondingly wide
  rural-urban gaps in chronic disease burden. Rural-targeted insurance enrollment efforts may offer the
   clearest opportunity for impact in Texas and New Mexico.

Oklahoma presents a different pattern and requires a different response. Its elevated chronic disease burden
 appears broadly across both urban and rural counties rather than concentrated in one, suggesting statewide
  programming may be more appropriate than a rural targeted approach.

As these suggested interventions are based on modeled prevalence data from surveys, they are best used to
 identify where to prioritize further investigation and resource planning, not as the sole evidence baseline for future actions.


### Appendix

The following tables provide additional statistical detail underlying the tests conducted in this analysis.

```{r,echo = FALSE}
analysis_objects$Table1%>%
  gt()%>%
  cols_label(Short_Question_Text = "Measure")%>%
  tab_source_note(source_note = md("Supplementary Table 1: Table of measure and p-value; for data shown in Figure 2."))%>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```
```{r, echo = FALSE}
analysis_objects$Table2%>%
  select(Short_Question_Text, effsize, magnitude)%>%
  gt() %>%
  cols_label(Short_Question_Text = "Measure", effsize = "Effect Size", magnitude = "Magnitude")%>%
  tab_source_note(source_note = md("Supplementary Table 2:Effect size and magnitude for data shown in Figure 2")) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```
```{r, echo = FALSE}
analysis_objects$Table5%>%
  gt()%>%
  cols_label(Short_Question_Text = "Measure", StateAbbr = "State", diff = "Difference")%>%
  tab_source_note(source_note = md("Supplementary Table 3: Detailed values associated with figure 5")) %>%
  tab_options(table.font.size = px(12),
    data_row.padding = px(3))
```